Detection of eco-evolutionary dynamics in metacommunities using Joint Species

### 2024-11-26 manuscript/manuscript_v13.Rmd
Updated S1, knit.

To fit linear regression models to the simulated population size data, we began with models using population size (model *A*) and environment (including the second-order term, to better reflect species optimum values; model *B*= as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$; for simulation model *E*) and time (models *C*, *D*, *E*, and *F*) and spatial distance (model *F*) as random effects. For consistency, all linear regression models were implemented as hierarchical Bayesian models using HMSC (Hierarchical Modelling of Species Communities [@Ovaskainen2017HMSC] [@Tikhonov2020]). HMSC uses a latent variable approach as an alternative to directly modelling species interaction coefficients, estimating a reduced number of linear combinations of species abundances that best predict future population size values. For linear models fit to data generated in simulations *C*, *D*, *E*, and *F* above, the time random effect was implemented to capture species interactions (as part of the temporal species association matrix estimated by the model), and the spatial random effect was implemented in model *F* to capture spatial covariance in abundances as a function of distance between sites. Full methodological details are given in Appendix S1. All simulations were run in R (@R2024) and HMSC models were implemented using the `Hmsc` package (@Hmsc2022).

@@

All simulations used the same parameter values ($W_{max}$ = 2, *P* = 1, *w* = 2, $N_c$ = 100, *p* = 0.001), species interaction matrix ($\alpha_{ii} = 0.00125$; $\alpha_{ij} \sim U(0.0015,1)$), and initial conditions (initial richness $s_{0k} \sim Pois(0.75)$; initial population size $N_{0i} \sim Pois(10)$; initial degree of maladaptation $\beta_{i0} \sim Gamma(0.75,1)$, initial distance to the local patch's optimum phenotype $d_{i0}=\sqrt{B_{i0}(w+P)}$, initial trait value $x_{i0} = E_k - d_{i0}$; Appendix S4). Simulations were run across a range of dispersal values (identical for all species; *d* = 0, $10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}$, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: $h^2$ = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) for 200 time steps, producing records of $N_{ikt}$ and $x_{ikt}$.

>>>>

To fit linear regression models to the simulated population size data, we began with models using population size (model *A*) and environment (including the second-order term, to better reflect species optimum values; model *B*) as fixed effects ($N_t \sim N_{t+1} + E_t + E_t^2$), and progressed to a mixed model using environment and trait evolution as fixed effects ($N_t \sim E_t + E_t^2 + |\Delta x|$; for simulation model *E*) and time (models *C*, *D*, *E*, and *F*) and spatial distance (model *F*) as random effects. For consistency, all linear regression models were implemented as hierarchical Bayesian models using HMSC (Hierarchical Modelling of Species Communities [@Ovaskainen2017HMSC] [@Tikhonov2020]). HMSC uses a latent variable approach as an alternative to directly modelling species interaction coefficients, estimating a reduced number of linear combinations of species abundances that best predict future population size values. For linear models fit to data generated in simulations *C*, *D*, *E*, and *F* above, the time random effect was implemented to capture species interactions (as part of the temporal species association matrix estimated by the model), and the spatial random effect was implemented in model *F* to capture spatial covariance in abundances as a function of distance between sites. Full methodological details are given in Appendix S1. All simulations were run in R (@R2024) and HMSC models were implemented using the `Hmsc` package (@Hmsc2022).

@@

All simulations used the same parameter values ($W_{max}$ = 2, *P* = 1, *w* = 2, $N_c$ = 100, *p* = 0.001), species interaction matrix ($\alpha_{ii} = 0.00125$; $\alpha_{ij} \sim U(0,0.0015)$), and initial conditions (initial richness $s_{0k} \sim Pois(0.75)$; initial population size $N_{0i} \sim Pois(10)$; initial degree of maladaptation $\beta_{i0} \sim Gamma(0.75,1)$, initial distance to the local patch's optimum phenotype $d_{i0}=\sqrt{B_{i0}(w+P)}$, initial trait value $x_{i0} = E_k - d_{i0}$; Appendix S4). Simulations were run across a range of dispersal values (identical for all species; *d* = 0, $10^{-9},10^{-8},10^{-7},10^{-6},10^{-5},10^{-4},10^{-3},10^{-2}$, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) and heritability values (identical for all species: $h^2$ = 0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1) for 200 time steps, producing records of $N_{ikt}$ and $x_{ikt}$.
